---
title: "Introduction: loading packages and downloading data"
teaching: 15
exercises: 5
---

:::::::::::::::::::::::::::::::::::::: questions

- How is whole genome methylation interrogated?
- How are methylation values commonly reported?
- What preliminary R packages and data is necessary to begin methylation analysis?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Setting up the R environment and required  packages
- Create a sample table and define the variable of interest
- Download the six publicly available sample data from GEO
- Import the IDAT files into a `minfi` object

::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

Methylation in the human genome is known to be associated with development and disease. The Illumina Infinium methylation arrays are by far the most common way to interrogate methylation across the human genome. This workshop provides an easy and reproducible workflow using multiple packages for the analysis of methylation array data. Specifically, we demonstrate the steps involved in a typical differential methylation analysis pipeline including: quality control, filtering, normalization, data exploration and statistical testing for probe-wise differential methylation.

DNA methylation, the addition of a methyl group to a CG dinucleotide of the DNA, is the most extensively studied epigenetic mark due to its role in both development and disease. Although DNA methylation can be measured in several ways, the epigenetics community has enthusiastically embraced the Illumina HumanMethylation450 (450k) array as a cost-effective way to assay methylation across the human genome. Illumina has increased the genomic coverage of the platform with their MethylationEPIC (850k) array, and more recently MethylationEPICv2 (935k) covering to >935,000 sites. As methylation arrays are likely to remain popular for measuring methylation for the foreseeable future, it is necessary to provide robust workflows for methylation array analysis.

For each CpG, there are two measurements: a methylated intensity (denoted by 𝑀) and an unmethylated intensity (denoted by 𝑈). These intensity values can be used to determine the proportion of methylation at each CpG locus. Methylation levels are commonly reported as either beta values (𝛽=𝑀/(𝑀+𝑈)) or M-values (𝑀𝑣𝑎𝑙𝑢𝑒=𝑙𝑜𝑔2(𝑀/𝑈)).

Beta values (proportion of methylation at a particular CpG site) are generally preferable for describing the level of methylation at a locus or for graphical presentation because percentage methylation is easily interpretable. M-values (log2 ratio of methylated to unmethylated signal intensity) are more appropriate for statistical testing.

## Load packages

Complete the [setup instructions](../learners/setup.md) first. Open RStudio and
create a project using **File > New Project > New Directory > New Project**,
or open your existing workshop project. Create an R script, and run the chunks below in order.

Installing a package downloads it to your computer. Loading it with `library()`
makes its functions available in the current session. Load the packages again
whenever you restart R.

```{r load-packages, eval=TRUE, echo=FALSE}
# Data handling
library(tidyverse)
library(readxl)
library(reshape2)

# Methylation analysis and annotation
library(minfi)
library(GEOquery)
library(limma)
library(IlluminaHumanMethylationEPICv2anno.20a1.hg38)
library(IlluminaHumanMethylationEPICv2manifest)
library(AnnotationHub)
library(DMRcate)
library(methylKit)
library(missMethyl)
library(clusterProfiler)
library(org.Hs.eg.db)

# Plotting and parallel computing
library(Polychrome)
library(ComplexHeatmap)
library(ggsci)
library(gridExtra)
library(RColorBrewer)
library(parallel)
```

## Load annotation data

This is information that describes each CpG probe on the Illumina EPICv2 methylation array. It tells you where each probe is located in the genome and which genes or genomic features it is associated with.

```{r load-annotation-data, eval=TRUE}
epic <- getAnnotation(IlluminaHumanMethylationEPICv2anno.20a1.hg38)
head(epic)
```

## Download data

For this tutorial we are using the publicly available LNCaP and PREC cell lines
(500 ng DNA input) from Peters *et al.* (2024), available from GEO
([GSE240469](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE240469)). This contains data for 3 technical replicates per cell line. 
The code below executes when this episode renders. It downloads the IDAT
files, imports the raw measurements, and saves the objects for the next episode.

If you would like to use your own data, create a sample table and variables of
interest in the same format, with your own IDAT files, samples and variables.
The GEO download chunk is only needed for data hosted on GEO.

### Create the sample table

Each row describes one sample. `Accession` identifies the GEO sample, `Name`
matches its filename label, and `Type` identifies the cell line to compare.

```{r sample-table, eval=TRUE}
# Create the sample table for IDAT downloads. Replace this table with your own sample information when using your own data.

targets <- data.frame(
  Accession = c(
    "GSM7698438",
    "GSM7698446",
    "GSM7698462",
    "GSM7698435",
    "GSM7698443",
    "GSM7698459"
  ),
  Name = c(
    "LNCAP_500_1",
    "LNCAP_500_2",
    "LNCAP_500_3",
    "PREC_500_1",
    "PREC_500_2",
    "PREC_500_3"
  ),
  Type = c("LNCaP", "LNCaP", "LNCaP", "PREC", "PREC", "PREC")
)

targets$Name2 <- paste(
  targets$Accession, targets$Name, sep = "_"
)

# Define the variable of interest and convert it to a factor.
variable.of.choice <- as.factor(targets$Type)

targets
```

### Download the IDAT files

The following chunk downloads the IDAT files for the six accessions in
`targets`. Each sample is saved in its own accession-named folder within
your current project directory.

```{r download-data, eval=TRUE}
# Download only IDAT files for the selected tutorial samples.
gsmlist <- lapply(
  targets$Accession,
  GEOquery::getGEOSuppFiles,
  makeDirectory = TRUE,
  baseDir = getwd(),
  filter_regex = "[.]idat([.]gz)?$"
)
names(gsmlist) <- targets$Accession
```

## Read in .idats

Add the sample names and file paths to `targets`, preserving its sample order.
`Basename` is the path to each file pair.

```{r read-idats, eval=TRUE}
# Link the sample information to the downloaded IDAT files.
targets$Sample <- targets$Name
targets$Basename <- file.path(
  getwd(), targets$Accession, targets$Name2
)

# Read in raw data from the red/green IDAT files.
rgSet <- read.metharray.exp(targets = targets)
sampleNames(rgSet) <- targets$Sample
rgSet@annotation <- c(
  array = "IlluminaHumanMethylationEPICv2",
  annotation = "20a1.hg38"
)
rgSet
```

`rgSet` contains the raw intensities for the six samples. For your own data,
set `targets$Basename` to your file paths and use the annotation matching your
array type. Quality control, filtering and normalisation follow in the next
episode.

## Save the input objects for reuse

`targets` contains sample information; `rgSet` contains the raw measurements.

```{r save-input-objects, eval=TRUE}
dir.create("data", showWarnings = FALSE)
saveRDS(targets, "data/targets.rds")
saveRDS(rgSet, "data/rgSet.rds")
```

::::::::::::::::::::::::::::::::::::: keypoints

- Raw Illumina methylation array data are stored in `.idat` files, with one red-channel and one green-channel file per sample.
- The `targets` data frame holds the sample metadata, linking each sample's name and cell-line group to its IDAT file paths.
- `read.metharray.exp(targets = targets)` reads the raw intensities into `rgSet`, an `RGChannelSet` object, ready for quality control and preprocessing.

::::::::::::::::::::::::::::::::::::::::::::::::
